A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-05-28 paper/paper.tex
paper/introduction: more polish, more content

For instance, in the case of a semi-cubical set, presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes, the standard presheaf approach prescribes a family of sets fibered the ones over the others (up to coherence conditions such). Formulated in type theory, it corresponds to:

@@

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{The title of the paper ``A dependently-typed construction of semi-simplicial types'' is confusing: it was part of a project at defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids).}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by~\cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case in the presheaf presentation.

\subsection*{Reynolds' parametricity and its unary and binary variants}
In the context of functional programming, Reynolds' parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, correspondences, in the sense of ``relevant'' relation, that is of a family over a product of types, ca Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\citep{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using families instead of relations or graphs, in which case, we obtain a form of realizability~\citep{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

\subsection*{Iterated-parametricity-based construction}
We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets, generated by iterating the parametricity translation for type judgements. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\citep{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

\subsection*{Contribution}
Our work:
\begin{itemize}
  \item characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions
  \item provides a new example of indexed constructions of semi-simplicial and semi-cubical sets~\citep{voevodsky12,herbelin15,part15}, useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\citep{part15,shulman14,altenkirch16,kraus2017space}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett})
  \item makes a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do; compare, for example, the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~(\citeyear{curien14}).
\end{itemize}
\subsection*{Outline}
The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

\section{Preliminaries}
Martin-Löf's Type theory~[\cite{martinlof75,martinlof84}] is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

@@

In Section~\ref{sec:ett} and Tables~\ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces}, \ref{tab:coh}, we give an informal, mathematical definition in Extensional Type Theory, which roughly corresponds to a level of reasoning ``à la set theory'' where equalities are thought as ``holding on the nose'' without having to mention them explicitly once they are justified. In Section~\ref{sec:itt} and Tables~\ref{fulltab:coind}, \ref{fulltab:core}, \ref{fulltab:frames}, \ref{fulltab:faces}, \ref{fulltab:coh}, we give the same mathematical definition in Intensional Type Theory, where transporting along coherence conditions is now made explicit. Adding this level of details helps to understand the specificities of coherence conditions.

>>>>

For instance, in the case of a semi-cubical set~\citep{grandis03,buchholtz17}, presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes, the standard presheaf approach prescribes a family of sets fibered the ones over the others (up to coherence conditions such). Formulated in type theory, it corresponds to:

@@

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{With hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somehow confusing: it was part of a project at defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids). The confusion was however common at that time.}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by~\cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case by default in the presheaf presentation. In category-theoretic terms, the indexation set of the family $X_n$ in the indexed presentation is called a \emph{matching object}. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

\subsection*{Reynolds' parametricity and its unary and binary variants}
In the context of functional programming, Reynolds' parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, correspondences, in the sense of ``relevant'' relation, that is of a family over a product of sets, can be used in place of relations. Parametricity can then be iterated, and, relying on the fibered representation of correspondences as spans, it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\citep{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or correspondences, in which case, we obtain a form of realizability~\citep{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch.}. This similarity suggests to join the definition of augmented semi-simplicial sets and cubical sets and to see both of them as instances of a more general construction which we call $\nu$-sets of presheaves over a $\nu$-semi-shape category made of words of some cardinal $\nu$+1, where $\nu=1$ gives augmented semi-simplicial sets and $\nu=2$ gives semi-cubical sets.

\subsection*{An iterated parametricity-based indexed construction}
The main contribution of the paper is to describe in full details a recipe uniformly characterizing unary and binary iterated parametricity in indexed form and to derive from it a new indexed presentation, called indexed \emph{$\nu$-sets}, uniformly of augmented semi-simplicial and semi-cubical sets. Alternatively, from a category-theory perspective, we are characterizing, among the class of matching objects up to isomorphism for augmented semi-simplicial and cubical sets, a particular instance that reflects the structure of iterated parametricity.

\subsection*{Contribution}
Our work:
\begin{itemize}
  \item Characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions.
  \item Provides a new example of indexed constructions of semi-simplicial and semi-cubical sets~\citep{voevodsky12,herbelin15,part15}, useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\citep{part15,shulman14,altenkirch16,kraus17}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}).
  \item Makes a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do; compare, for example, the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~(\citeyear{curien14}).
\end{itemize}
\subsection*{Outline}
The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

\section{Preliminaries}
Martin-Löf's Type theory~\citep{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~(\citeyear{agda23}), Coq~(\citeyear{coq22}) or Lean~(\citeyear{lean15}). In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

@@

In Section~\ref{sec:ett} and Tables~\ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces}, \ref{tab:coh}, we give an informal, mathematical definition in Extensional Type Theory, which roughly corresponds to a level of reasoning ``à la set theory'' where equalities are thought as ``holding on the nose'' without having to mention them explicitly once they are justified. In Section~\ref{sec:itt}, we give the same mathematical definition in Intensional Type Theory, where transporting along coherence conditions is now made explicit. Adding this level of details helps to understand the specificities of coherence conditions.
